Crop diversification in a game against nature in a mediterranean agro-ecosystem

### 2020-01-23 risk-percption.Rmd
pure strategy plots

||Dry Year|Normal Year|Wet Year|
|-------|-----|-----|----|----|
|Barley Yield|0.93|1.18||
|Wheat Yield|0|1.60||

@@

```{r}
wheat_dry <- 0
wheat_normal <- 1.5
wheat_wet <- 1.7

barley_dry <- 0.8
barley_normal <- 1
barley_wet <- 0.5
```

@@

* Maximax
2. Maximize *subjective* expected utility
3. Satisficing learning -- win-stay lose-move
4. Nash equilibrium -- minimax (risk free)
5. Fictitious play learning rule -- use empirical frequency distribution.
A. Good against environment when environment is stationary but not when its nonstationary.
B. Bayesian belief based learning

The pure maximin strategy is to plant barley:
```{r}
max(wheat_dry, barley_dry)
```

@@

Given these payoffs, what's the point of indifference between the two crop types?
```{r, echo = FALSE}
indiff <- tibble(risk = (barley_wet - wheat_wet) / (wheat_dry - wheat_wet - barley_dry + barley_wet),
                 yield = risk * wheat_dry + (1 - risk) * wheat_wet)
indiff_label <- paste0('(', round(indiff$risk, 2), ', ', round(indiff$yield, 2), ')')

tibble(risk = seq(0, 1, .1),
      barley_exp = risk * barley_dry + (1 - risk) * barley_wet,
      wheat_exp = risk * wheat_dry + (1 - risk) * wheat_wet) %>%
  gather(type, yield, barley_exp:wheat_exp) %>%
  ggplot(aes(risk, yield)) +
  geom_hline(yintercept = indiff$yield, linetype = 2)+
  geom_vline(xintercept = indiff$risk, linetype = 2) +
  geom_line(aes(color = type), size = 1.2) +
  geom_point(data = indiff, color = 'black', size = 3) +
  geom_text(data = indiff, label = indiff_label, nudge_x = 0.1, nudge_y = 0.1) +
  theme_classic() +
  ggtitle('Point of indifference between wheat and barley', 'Given varying drought risk')
```

The expected yield is thus calculated as.
```{r}
expected_yield_wheat <- function(p_dry, p_wet) {
  p_dry * wheat_dry + p_wet * wheat_wet + (1 - p_wet - p_dry) * wheat_normal 
}

expected_yield_barley <- function(p_dry, p_wet) {
  p_dry * barley_dry + p_wet * barley_wet + (1 - p_wet - p_dry) * barley_normal 
}
```

```{r}
expected_yield_wheat(p_dry = 0.3, p_wet = 0.2)
expected_yield_barley(p_dry = 0.3, p_wet = 0.2)
```

```{r}
expand_grid(wheat_prop = seq(0, 1, .01),
       risk = seq(.1, .9, .1)) %>%
  mutate(wheat_exp = wheat_prop * expected_yield_wheat(risk,0),
      barley_exp = (1 - wheat_prop) * expected_yield_barley(risk, 0),
      exp_yield = wheat_exp + barley_exp) %>%
  ggplot(aes(wheat_prop, exp_yield, group = risk, color = risk)) +
  geom_line()
```

The risk averse MAXIMIN pure strategy is to plant barley all the time, and the risk indifferent farmer plants wheat for its high yields

The maximin strategy here is to plant barley all the time, but this entails a huge opportunity cost over planting wheat or a mix of the two. So the intersection point of the next plot is the mixed strategy equilbrium (indifference point) in a risk-free environment

```{r, echo = FALSE}
nash_eq <- tibble(wheat_prop = round((barley_wet - barley_dry) / (wheat_dry + barley_wet - (wheat_wet + barley_dry)), 2),
                  utility = round(wheat_wet * wheat_prop + barley_wet * (1 - wheat_prop), 2))

tibble(wheat_prop = seq(0, 1, .1),
       dry_exp = wheat_prop * wheat_dry + (1 - wheat_prop) * barley_dry,
       normal_exp = wheat_prop * wheat_normal + (1 - wheat_prop) * barley_normal,
      wet_exp = wheat_prop * wheat_wet + (1 - wheat_prop) * barley_wet) %>%
  gather(type, utility, dry_exp:wet_exp) %>%
  ggplot(aes(wheat_prop, utility)) +
  geom_hline(yintercept = nash_eq$utility, linetype = 2)+
  geom_vline(xintercept = nash_eq$wheat_prop, linetype = 2) +
  geom_line(aes(color = type), size = 1.2) +
  geom_point(data = nash_eq, color = 'black', size = 3) +
  geom_text(data = nash_eq, 
            label = paste0('(', nash_eq$wheat_prop, ', ', nash_eq$utility, ')'),
            nudge_x = 0.1, nudge_y = 0.05) +
  theme_classic()
```

```{r}
# farmer's ne
pmax(0, round((barley_wet - barley_dry) / (wheat_dry + barley_wet - (wheat_wet + barley_dry)), 2))
#nature's ne
round((barley_wet - wheat_wet) / (wheat_dry + barley_wet - (wheat_wet + barley_dry)), 2)
```

@@

```{r}
learn <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_wet; barley_yield <- barley_wet

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_utility(risk),
         barley_exp = risk * barley_dry + (1 - risk) * barley_wet,
         wheat_exp = risk * wheat_dry + (1 - risk) * wheat_wet,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = drought + alpha,
           beta = beta + 1 - drought,
           mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}
```

@@

@@

simulation2 <- accumulate(droughts, ~learn_satisficing(.x, .y, aspiration = .8), .init = test3) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)

```{r}
simulation %>%
  group_by(time) %>%
  summarise(famine = sum(payoff == 0)) %>%
ggplot(aes(time, famine)) +
  geom_line() +
  theme_classic()
```

>>>>

||Dry Year|Normal Year|
|-------|-----|-----|----|
|Barley Yield|0.93|1.18|
|Wheat Yield|0|1.60|

@@

```{r}
wheat_dry <- 0
wheat_normal <- 1.6

barley_dry <- 0.93
barley_normal <- 1.18
```

@@

The pure maximin strategy is to plant barley:

@@

The expected yield is thus calculated as.
```{r}
expected_yield_wheat <- function(risk) {
  risk * wheat_dry + (1 - risk) * wheat_normal
}

expected_yield_barley <- function(risk) {
  risk * barley_dry + (1 - risk) * barley_normal
}

expected_yield <- function(risk) {
  expected_yield_wheat(risk) + expected_yield_barley(risk)
}
```

Given these payoffs, what's the point of indifference between the two crop types?
```{r, echo = FALSE, fig.cap=title = 'Expected wheat and barley yields under increasing drought risk with the point of indifference highlighted.'}
indiff <- tibble(risk = (barley_normal - wheat_normal) / (wheat_dry - wheat_normal - barley_dry + barley_normal),
                 yield = risk * wheat_dry + (1 - risk) * wheat_normal)
indiff_label <- paste0('(', round(indiff$risk, 2), ', ', round(indiff$yield, 2), ')')

tibble(risk = seq(0, 1, .1),
      Barley = expected_yield_barley(risk),
      Wheat = expected_yield_wheat(risk)) %>%
  gather(type, yield, Barley:Wheat) %>%
  ggplot(aes(risk, yield)) +
  geom_hline(yintercept = indiff$yield, linetype = 2)+
  geom_vline(xintercept = indiff$risk, linetype = 2) +
  geom_line(aes(color = type), size = 1.2) +
  geom_point(data = indiff, color = 'black', size = 3) +
  geom_text(data = indiff, label = indiff_label, nudge_x = 0.1, nudge_y = 0.1) +
  theme_classic() +
  scale_color_discrete(name = 'Crop') +
  labs(x = 'Drought probability', y = 'Expected yield')
```

```{r}
# farmer's ne
pmax(0, round((barley_normal - barley_dry) / (wheat_dry + barley_normal - (wheat_normal + barley_dry)), 2))
#nature's ne
round((barley_normal - wheat_normal) / (wheat_dry + barley_normal - (wheat_normal + barley_dry)), 2)
```

@@

```{r}
learn <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = drought + alpha,
           beta = beta + 1 - drought,
           mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}
```

@@

```{r}
simulation %>%
  group_by(time) %>%
  summarise(famine = sum(payoff < .1)) %>%
ggplot(aes(time, famine)) +
  geom_line() +
  theme_classic()
```

```{r}
simulation %>%
  group_by(time) %>%
  summarise(payoff = mean(payoff)) %>%
ggplot(aes(time,payoff)) +
  geom_point(aes(color = time)) +
  theme_classic()
```

```{r}
library(ggridges)
simulation %>%
  mutate(time_step = 5 * round(time/5),
         famine = payoff == 0) %>%
  ggplot(aes(famine, as.factor(time_step))) +
  geom_density_ridges(rel_min_height = 0.005) +
  theme_ridges()
```

make this a function to calculate payoffs for all wheat or all barley strategies as a comparison
```{r}
tibble(drought = droughts,
       barley = if_else(droughts == FALSE, barley_normal, barley_dry),
       wheat = if_else(droughts == FALSE, wheat_normal, wheat_dry),
       equal = (barley + wheat) / 2) %>%
  mutate(time = 1:n()) %>%
  gather(crop, yield, barley:equal) %>%
  ggplot(aes(time, yield, group = crop, color = crop)) +
  geom_point()
```

```{r}
tibble(drought = droughts,
       barley = if_else(droughts == FALSE, barley_normal, barley_dry),
       wheat = if_else(droughts == FALSE, wheat_normal, wheat_dry),
       equal = (barley + wheat) / 2) %>%
  mutate(time = 1:n()) %>%
  gather(crop, yield, barley:equal) %>%
  mutate(famine = yield == 0) %>%
  ggplot(aes(time, famine, group = crop, color = crop)) +
  geom_line()
```

simulation2 <- accumulate(droughts, ~learn_satisficing(.x, .y, aspiration = .5), .init = test3) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)

@@
